Subglacial and subaerial fluvial sediment transport capacity respond differently to water discharge variations

### 2024-07-01 Delaneyetal2024.tex
version to pass around

The second numerical experiment aims to quantify the sources of increased variability in the subglacial model outputs as they exhibit different seasonal evolutions and peaks (Figure~\ref{fig:model_outs}).
Variable relationships between model outputs and water discharge emerge for the subaerial and subglacial cases due to the hysteresis in channel size in the subglacial model (Equation~\ref{eq:dS_dt}).
Because subaerial channels have no history dependence, each water discharge value in the subaerial channel produces a unique water velocity, shear stress, and width-integrated shear stress (Section~\ref{sect:sub_mode}).
This characteristic results in a perfect rank correlation between the variables and water discharge (Figure~\ref{fig:Qw_vari}).
\begin{figure}[hbt!]
  \centering
  \includegraphics[width=0.9\linewidth]{Fig5.png}
  \caption{Model outputs from simulations using the hydrographs in panels a and e for the scenarios \alpine{} (a-d) and \icesheet{} (e-h).
    Black (gray) lines in a and e represent the subglacial channel size (water discharge).
    Blue (orange) lines represent outputs from the subglacial (subaerial) channel.
    Data are shown at $15$\unit{min} intervals.
    Arrows show examples where variables peak in the subglacial channel before the subaerial one.
    Insets in f-h show the peak melt event denoted by the shaded area in panels e--h, with an arbitrary y-axis.
  }
  \label{fig:model_outs}
\end{figure}

@@

Several aspects of the models and experiment design limit make comparing the model outputs to field observations difficult and make comparison of the two systems difficult.
The lumped nature of the models means that they operate independently of the upstream drainage network.
Additionally, sediment access is not considered here.
In reality, processes such as sediment access are highly important in controlling sediment export in glacierized catchments \citep[e.g.][]{herman2015,vergara2022}.
Furthermore, meltwater can be distributed to flow through several adjacent conduits impacting the sediment transport capacity in each \citep[e.g.][]{werder2013,hewitt2019,delaney2023}.
The lumped models used here make the comparison more direct by isolating the relationship between water discharge and sediment transport capacity in subglacial and subaerial systems. 
However, they neglect more complex, yet important, spatially distributed processes.

@@

Catchments with reduced water discharge variability may experience yet less variability in sediment transport capacity, shown by the \alpine{} and \icesheet{} model outputs.
Indeed, the more decoupled and sporadic relationship between model outputs and water discharge in \alpine{} results from the relatively larger discharge variations on sub-daily to weekly timescales (Figure~\ref{fig:Qw_vari}).
This occurs as the subglacial channel's size evolves slowly compared to the variations in water discharge.
High water discharge variability in \alpine{} may cause shear stress to approach $Q^{2}$in assuming pipe-flow conditions where the water discharge varies largely compared to channel size \citep[Section \ref{sect:scaling}; c.f.][]{alley1997}.
Conversely, the reduced variability in water discharge in \icesheet{} may result in the subglacial channel's size being closer to equilibrium from the smaller variations in water discharge.
In this case, the exponent on water discharge for shear stress is likely substantially less than $2$, but greater than $\frac{4}{5}$ that occurs in a steady-state R-channel (see Section \ref{sect:scaling}).
As a result, there is a stronger relationship between subglacial model outputs and water discharge in \icesheet{} (Figure~\ref{fig:Qw_vari}) and less variability model outputs in the \icesheet case (Figure~\ref{fig:multi_run}).

@@

Concerning the manuscript's first objective,
results demonstrate that sediment transport capacity variability is higher in subglacial channels with evolving size across a range of channel shapes and friction factors.
Increased variability occurs even in environments where the water discharge has relatively small diurnal variations in the \icesheet case.
Yet, elevated variability is especially strong with highly variable water discharge in the \alpine case.
Increased noise in sediment transport capacity in evaluating hydro-climatic signals from sediment records in glacierized catchments needs further evaluation.
Furthermore, this process may lead to sediment exhaustion by subglacial water repeatedly crossing the mobilization threshold across a range of water discharges.

The manuscript's second objective is to establish if water discharge varies with subglacial sediment transport capacity.
Results here suggest that, even in a transport-limited subglacial system, in most systems, an incoherent relationship between water and subglacial sediment discharge could be expected.
This presents a challenge in linking hydro-climatic conditions or events to sediment export from glaciers.
Due to evolving subglacial channel size, the timing of peak sediment discharge capacity in subglacial channels occurs before peak water discharge during an event, whereas it is coincident in subaerial ones.
Our analysis shows, however, that covariance between water discharge and sediment transport capacity could be possible when water discharge varies at a slower rate than subglacial channel size.
In natural environments, this appears to often not be the case, except where basal melt dominates.

>>>>

The second numerical experiment aims to quantify the sources of increased variability in the subglacial model outputs as they exhibit different seasonal evolutions and peaks (Figure~\ref{fig:model_outs}).
Variable relationships between model outputs and water discharge emerge for the subaerial and subglacial cases due to the hysteresis in channel size in the subglacial model (Equation~\ref{eq:dS_dt}).
Because subaerial channels have no history dependence, each water discharge value in the subaerial channel produces a unique water velocity, shear stress, and width-integrated shear stress (Section~\ref{sect:sub_mode}).
This characteristic results in a perfect rank correlation between the variables and water discharge (Figure~\ref{fig:Qw_vari}).
An inconsistent relationship between subglacial model outputs and water discharge persists across the range of parameters examined in the ensemble runs (Figure S15).
\begin{figure}[hbt!]
  \centering
  \includegraphics[width=0.9\linewidth]{Fig5.png}
  \caption{Model outputs from simulations using the hydrographs in panels a and e for the scenarios \alpine{} (a-d) and \icesheet{} (e-h).
    Black (gray) lines in a and e represent the subglacial channel size (water discharge).
    Blue (orange) lines represent outputs from the subglacial (subaerial) channel.
    Data are shown at $15$\unit{min} intervals.
    Arrows show examples where variables peak in the subglacial channel before the subaerial one.
    Insets in f-h show the peak melt event denoted by the shaded area in panels e--h, with an arbitrary y-axis.
  }
  \label{fig:model_outs}
\end{figure}

@@

Several aspects of the models and experiment design limit make comparing the model outputs to field observations difficult and make comparison of the two systems difficult.
The lumped nature of the models means that they operate independently of the upstream drainage network.
Additionally, sediment access is not considered here.
In reality, processes such as sediment access are highly important in controlling sediment export in glacierized catchments \citep[e.g.][]{herman2015,vergara2022}.
Furthermore, meltwater can be distributed to flow through several adjacent conduits impacting the sediment transport capacity in each \citep[e.g.][]{werder2013,hewitt2019,delaney2023}.
The lumped models used here make the comparison more direct by isolating the relationship between water discharge and sediment transport capacity in subglacial and subaerial systems. 
However, they neglect more complex, yet important, spatially distributed processes. a

@@

Catchments with reduced water discharge variability may experience yet less variability in sediment transport capacity, shown by the \alpine{} and \icesheet{} model outputs.
Indeed, the more decoupled and sporadic relationship between model outputs and water discharge in \alpine{} results from the relatively larger discharge variations on sub-daily to weekly timescales (Figure~\ref{fig:Qw_vari}).
This occurs as the subglacial channel's size evolves slowly compared to the variations in water discharge.
High water discharge variability in \alpine{} may cause shear stress to approach $Q^{2}$in assuming pipe-flow conditions where the water discharge varies largely compared to channel size \citep[Section \ref{sect:scaling}; c.f.][]{alley1997}.
Conversely, the reduced variability in water discharge in \icesheet{} may result in the subglacial channel's size being closer to equilibrium from the smaller variations in water discharge.
In this case, the exponent on water discharge for shear stress is likely substantially less than $w\tau \propto Q^2$, but greater than $w \tau \propto Q^{\frac{4}{5}}$ that occurs in a steady-state R-channel (see Section \ref{sect:scaling}).
As a result, there is a stronger relationship between subglacial model outputs and water discharge in \icesheet{} (Figure~\ref{fig:Qw_vari}) and less variability model outputs in the \icesheet case (Figure~\ref{fig:multi_run}).

@@

Concerning the manuscript's first objective,
results demonstrate that sediment transport capacity variability is high in subglacial channels with evolving size across a range of channel shapes and friction factors.
Increased variability occurs even in environments where the water discharge has relatively small diurnal variations in the \icesheet case.
Yet, elevated variability is especially strong with highly variable water discharge in the \alpine case.
Increased noise in sediment transport capacity in evaluating hydro-climatic signals from sediment records in glacierized catchments needs further evaluation.
Furthermore, this process may lead to sediment exhaustion by subglacial water repeatedly crossing the mobilization threshold across a range of water discharges.

The manuscript's second objective is to establish if water discharge varies with subglacial sediment transport capacity.
Results here suggest that, even in a transport-limited subglacial system, in most systems, an incoherent relationship between water and subglacial sediment discharge could be expected.
This presents a challenge in linking hydro-climatic conditions or events to sediment export from glaciers.
Due to evolving subglacial channel size, the timing of peak sediment discharge capacity in subglacial channels occurs before peak water discharge during an event, whereas it is coincident in subaerial ones.
Covariance between water discharge and sediment transport capacity could be possible when water discharge varies at a slower rate than subglacial channel size.
In natural environments, this appears to often not be the case, except where basal melt dominates.
